Searching for circuit complexity lower bounds using SAT solvers

### 2022-05-14 report/draft1.tex
report writing

\begin{titlepage}
\title{Searching for circuit complexity lower bounds using SAT solvers}
\maketitle
\end{titlepage}

\begin{abstract}
Where the abstract will go
\end{abstract}

@@

Finding lower bounds on the size of Boolean circuits computing given Boolean functions is one of the most fundamental and most challenging problems in computer science. Despite the fact that the majority of functions require exponential circuits, the best lower bounds proven on unrestricted circuits are only linear~\cite{boppana} and we still do not have optimal circuits or close upper and lower bounds for many important functions. Results in this area can be used to separate computational complexity classes, possibly even P and NP (i.e. to show that a large class of fundamental and practically useful problems do not have polynomially fast algorithms).~\cite{arora}

If functions with circuits of a given complexity could be identified by an efficient algorithm, this could be used to construct efficient learning algorithms for circuits of a smaller complexity; this would resolve one of the biggest questions in learning theory.~\cite{carmosino} Furthermore, such an algorithm could be used to break pseudorandom generators (i.e. distinguish random from pseudorandom inputs).~\cite{arora} As a result, this problem has strong connections in both learning theory and cryptography.

A heuristic approach is to find minimal circuits for small instances of the functions. This can be used to prove new bounds~\cite{kulikovsurvey} and may lead to theoretical insights on the structure of their optimal circuits generally~\cite{williams}. Finding efficient circuits is also necessary in practice when designing electronic systems.

One approach to automating the search for efficient circuits is to reduce the problem of designing a correct circuit (logical design synthesis) to the Boolean satisfiability problem (SAT). Existing algorithms (SAT solvers) can then be used to solve the problem, either finding a correct circuit of fixed size or generating a proof that none exists~\cite{kulikov}. SAT solvers have become increasingly powerful in recent years *cite* and now have many industrial applications*cite*; the wide range of heuristics implemented by different SAT solvers means that one suited to the problem at hand could be found or developed. *cite*

*how to extend it*

@@

Circuits are a popular model of non-uniform computation. In the past there have been theoretical proofs of lower bounds for restricted circuit classes, e.g. AC0\cite{furst81}\cite{ajtai83}, AC0[k] (i.e. AC0 additionally with MOD-k gates)\cite{razborov87}\cite{smolensky87}, and monotone circuits\cite{razborov85}. However, no lower bounds better than 5n-c\cite{iwama02} have been found for general circuits. Furthermore, there are results stating that it is impossible to use circuit lower bounds to separate P and NP by relativising methods\cite{baker75}, algebrizing methods\cite{aaronson09}, or (under standard crpytpgraphic assumptions) so-called 'natural proofs'\cite{razborov94}. It has been proved that circuit lower bounds are hard to prove using resolution and resolution with k-DNFS, and it is conjectured that they are also hard under stronger proof systems such as Frege systems\cite{raz}. Due to this difficulty in theoretical progress, we would like to investigate the problem empirically.

Kamath et al~\cite{kamath} propose a reduction from logical design synthesis to SAT which fixes the DNF structure of the circuit and number of disjunctions used. Experiments using this reduction to find minimal circuits for 2-4 bit adders and multipliers are reported in~\cite{estrada}. Kojevnikov et al~\cite{kulikov} demonstrate a more general reduction allowing any circuit with a fixed number of gates, proving linear bounds for \(MOD^n_3\) circuits over different Boolean bases. This was done by checking for circuit 'building blocks' with 5 to 11 gates. In~\cite{kulikovlocal} Kulikov et al. used the same reduction to improve large circuit building blocks by searching for smaller versions of their sub-circuits. Specifically, linear upper bounds were proved or verified for \(SUM\), \(MOD^n_4\), and \(MOD^n_3\) by checking for circuits with up to 13 gates (they remarked that checking for existence of a Boolean circuit of size 13 was out of reach). Knuth also used SAT solvers to investigate lower bounds for \(MOD^n_3\) \cite{knuth15}.

@@

We also search for a general relationship between solver performance and minimum circuit size by applying our methods to a sample of randomly generated functions. We then use solvers to prove a result, that a lower bound conjecture about MOD-3 functions due to Knuth\cite{knuth15} does not hold for the de Morgan basis.

@@

Denote by \(B_{n,m}\) the set of all Boolean functions \(f: \{0,1\}^n \to \{0,1\}^m\) and let \(B_n = B_{n,1}\). A circuit over the basis \(A \subseteq B_2\) is a directed acyclic graph with nodes of in-degree 0 or 2. Nodes of in-degree 0 are called inputs. Nodes of in-degree 2 are assigned functions from \(A\) and are called gates. Some nodes may be also be marked as outputs.

Without loss of generality, we can also assert that every gate has a larger index than both of its predecessors (i.e. the gates are sorted topologically w.r.t. the used numbering); that every gate's second predecessor has a larger index than its first predecessor; and that the last gate is an output.~\cite{kulikov}

A circuit \(c\) of size N is said to compute the function \(f\in B_{n,m}\) if, for all input vectors \(v\in\{0,1\}^n\) in the range of \(f\), \(c(v)=f(v)\), where \(c(v) = (o_1,...o_m)\) are the values of the output gates of \(c\), each input \(x_i\) takes value \(v_i\), and each gate takes the value of its assigned function when applied to the value of its two predecessor nodes.

In this paper, we mainly consider circuits over the De Morgan basis \[C_2 = \{\lor, \land, \neg\}\] where \(\neg\) denotes the Boolean function which outputs the negation of its first argument.

The size of a circuit is its number of gates. We define the circuit existence problem as the question of whether, given a truth table for a function \(f: \{0,1\}^n \to \{0,1\}^m\), there exists a circuit of size N computing \(f\).

@@

SAT solvers have previously seen success in proving mathematical theorems, most famously in 2016 when Heule et al solved the Boolean Pythagorean Triples problem\cite{heule} and in 2020 when Brakensiek, Heule et al solved Keller's conjecture\cite{heule2}. In general, SAT solvers have the advantage of being effective even when heuristics are not known for solving the original problem, as in the case of circuit existence.

We used three SAT solvers, all based on conflict-driven clause learning. This is an algorithm closely related to DPLL, and which also makes use of the resolution proof system.\cite{cdcl}\cite{krajicek} As mentioned earlier, this means that they cannot decide general circuit lower bounds with a less than exponential-length proof.\cite{raz} However, this asymptotic bound still allows for reasonably-sized proofs for small inputs such as the ones investigated in this paper.

The general idea of CDCL is to choose an assignment to a variable, propagate it (simplify the problem by removing clauses that now evaluate to true), and repeat until either finding a satisfying assignment, or finding that our assignment makes the formula unsatisfiable: in which case, add a 'learned clause' to the formula according to a procedure incorporating resolution, and then 'backtrack' (remove recent assignments so the formula is no longer unsatisfiable - if this is impossible, the answer is UNSAT) before trying again.

MiniSAT\cite{minisat} was the winner of the 2005 SAT Competition and was last updated to version 2.2 in 2010\cite{minisat2010}. Despite being old, it is well-documented and the other solvers we used were based on it, making it a good baseline for performance. It was also one of the solvers used to decide circuit existence in \cite{kulikov}.

PicoSAT is a solver inspired by miniSAT 1.14, with a focus on improving low-level performance by using optimised data structures\cite{picosat}. It was submitted in the 2010 SAT-Race, and was also one of the solvers used in \cite{kulikov} and the main one in \cite{kulikovlocal}.

@@

Finally, in order to make full use of the Oxford ARC service and its distributed computing, the parallel solver Glucose-Syrup was used. This is a 2014 parallel version of the Glucose solver, which placed 4th in both the 2014 and 2015 SAT Competition. Glucose is a solver also based on MiniSAT 2.2, which uses a new heuristic to more accurately estimate the 'usefulness' of its learned clauses and regularly delete ones that are not as useful \cite{glucose}. Glucose-Syrup is an unusual 'non-portfolio' parallel solver which runs multiple instances of the same solver in parallel, but rather than taking a divide and conquer approach, uses new heuristics for sharing learned clauses between the instances in order to speed them up \cite{glucose-syrup}.

@@

\begin{enumerate}
  \item Let $l_1$, $l_2$ be existing literals chosen from the formula and let q be a fresh variable (the extension variable).
  \item Add the clauses
  \[
    {\neg l_1, q}, {\neg l_2, q}, {l_1, l_2, \neg q}
  \]
  These are known as extension axioms.
  \item Repeat arbitrarily many times with new choices for $l_1$, $l_2$ and new fresh variable $q$. Note that the previously added q may also be chosen as a literal.
\end{enumerate}

These axioms allow the SAT solver to use 'extended resolution' - resolving on a formula incorporating multiple literals, rather than on a single literal. This is possible since $q \leftrightarrow l_1 \lr l_2$ and $l_1$, $l_2$ may also be extension variables, allowing for larger formulae to be built. \cite{krajicek}

@@

We encoded the circuit existence problem as a Boolean formula using three reductions. Let $t$ be the target function with $n$ inputs, and let $N$ be the number of gates required in the circuit.

@@

The first is due to Kojevnikov et al~\cite{kulikov} who present it together with bounds on the growth rate of the formula. In particular, the growth rate of the number of clauses is $O(N \cdot 2^n)$. The basis of the circuit can be specified as any subset of \(B_2\).

@@

The second is due to Razborov~\cite{raz}, assumes the function outputs only a single Boolean value, and requires a circuit over the basis \(C_2\). Note that it has a lower growth rate (in terms of number of clauses) than the other two reductions - $O(N \cdot 2^n)$ as opposed to $O(N^3 \cdot 2^n)$ - suggesting it encodes the requirements more efficiently.

@@

The third reduction is straightforward. It also assumes the function outputs only a single Boolean value and requires a circuit over the basis \(C_2\). The growth rate for the number of clauses is $O(N \cdot 2^n)$, the same as for the Kojevnikov reduction.

@@

Timed experiments were carried out on a 3.400GHz Intel i3-8130U processor running Linux. The times shown were obtained by running the software on the problem 3 times and taking the median.

Further experiments were conducted on the Oxford ARC service *cite* using 1 node which could run up to 32 processes in parallel. *add more details about ARC pleas

@@

We investigated the relationship between the size of a circuit and the time taken for a SAT solver to confirm or deny its existence. We encoded a range of simple symmetric functions using the Kulikov encoding:

@@

The size of the minimal circuit is shown in red*need to re-plot the graphs to make this work better*. It was found that the time needed to solve a problem tended to spike on or around the optimal circuit size. This may be because it is difficult to confirm or rule out circuits that come close to satisfying the requirements. Another possibility is that when close to the optimal size, there is the lowest proportion of satisfying circuits relative to the total amount of possible circuits, meaning that more of the search space must be explored (as opposed to quickly exploring it all at small sizes or quickly landing upon a satisfying circuit at large sizes). A similar phenomenon was reported in~\cite{estrada}.

Different functions took different amounts of time overall, varying between seconds and hours, something also seen in Section \ref{randomresults}. Performance for each function was not especially consistent between the solvers; usually they found the same things difficult, but the length of time each solver took for the difficult problems varied, for example in \ref{fig:8andkulikov} or. The extent of this can be seen in Section \ref{satresults}.

@@

\begin{figure}[!ht]
  \includegraphics[width=\textwidth]{images/times/8andkulikov.png}
  \label{fig:8andkulikov}
  \caption{Times taken by SAT solvers to check for existence of a circuit computing AND on 8 bits}
\end{figure}

\begin{figure}[!ht]
  \includegraphics[width=\textwidth]{images/times/4paritykulikov.png}
  \label{fig:4paritykulikov}
  \caption{Times taken by SAT solvers to check for existence of a circuit computing parity on 4 bits}
\end{figure}

@@

We investigated whether there was a relationship between the reduction used to encode a Boolean function, the solver used, and the time taken by the solver. We took the total of the times each combination took to solve the problems from Section \ref{funcresults} and compared them as shown in \ref{totals}.

\begin{figure}[!ht]
  \includegraphics[width=\textwidth]{images/totals.png}
  \label{fig:totals}
  \caption{Comparison of times taken by SAT solvers on different reductions}
\end{figure}

It can be seen that MapleSAT on the Kulikov encoding performs the best. Apart from this, the effect of the reduction and the solver seem to be independent, with MapleSAT consistently performing better than MiniSAT and PicoSAT, and the Naive and Kojevnikov reductions performing better than the Razborov reduction. It is possible that the efficient encoding of the Razborov reduction is detrimental to the performance of the solver. However, these results are only applicable to small values of n.

@@

\begin{figure}[!ht]
  \includegraphics[width=\textwidth]{images/extvars/boxplot_4parity_medians.png}
  \label{fig:extparity}
  \caption{Comparison of times taken to check existence of a 9-gate circuit for 4-bit parity}
\end{figure}

\begin{figure}[!ht]
  \includegraphics[width=\textwidth]{images/extvars/boxplot_4mod3_medians.png}
  \label{fig:extmod3}
  \caption{Comparison of times taken to check existence of a 9-gate circuit for \(MOD^0_3\)}
\end{figure}

@@

@@

The scatter graph shows a clustering in times between functions of the same optimal circuit size, but with a few outliers taking more time. This is most easily seen for sizes 10 and 11 (although the same is visible for 8 and 9 when plotted on a smaller scale). Also, the clustered functions of each optimal circuit size took strictly less time to solve than those with larger sizes, despite the fact that we always checked for existence up to 11 gates. This shows a generally strong relationship between the difficulty of checking circuit existence and the complexity of the function.

While we expect a large proportion of circuits to be hard (i.e. have a large optimal circuit size) for large input sizes~\cite{arora}, the usual counting arguments break down for such small values of n. However, using the SAT solver we can check the number of hard circuits directly. While the small sample size can account for the irregular shape of the graph in \ref{fig:4rand_sizes}, there does appear to be a peak around 10 gates. This contradicts the intuition that there should be exponentially more harder functions, and suggests that there is quite a low upper limit on the complexity of a function which only can only vary over 4 bits of input. However, this is all under the assumption that the pseudorandom generator used did output a truly random sample; in reality, it may have sampled disproportionately many easy (or hard) functions.

\begin{figure}[!ht]
  \includegraphics[width=\textwidth]{images/random/4bit_scatter.png}
  \label{fig:4rand_scatter}
  \caption{Time taken to check up to 11 gates for randomly chosen functions}
\end{figure}
\begin{figure}[!ht]
  \includegraphics[width=\textwidth]{images/random/4bit_sizes.png}
  \label{fig:4rand_sizes}
  \caption{Sizes of optimal circuits for 4-bit functions with randomly chosen truth tables}

\end{figure}

@@

Define $MOD^{m,r}_n(x_1 \ldots x_n) = [x_1 + \lodts + x_n \equiv r $mod$ m]$, where $[b]$ is the Iverson bracket, i.e. evaluates to 1 if $b$ evaluates to true, else false.

@@

We decided to try and check the hypothesis for $n=6, r=0$ and $n=6, r=1$ for the De Morgan basis. This involved checking for circuits for the function up to size 11 and 12, and 12 and 13 respectively. The Glucose-Syrup solver was used, running on the Oxford ARC service, which displayed gains in performance as the number of threads (so, parallel instances of the server working together) increased up to 32. *need to include some specs of ARC and citation*

@@

*honestly, not sure how to draw conclusions from this or whether I can add a few more results*

@@

The main practical lesson learned from this project was to think in the long term. Firstly, by writing code that can easily be adapted, i.e. following good OOP practices, so that less refactoring will be needed later, and by writing code that can scale to dealing with larger numbers as the scope and ambition of the project expands. This could be done by choosing a well-suited programming language, algorithms and data structures - I chose TypeScript and used arrays to store my CNF formulae, since these were familiar and easy for me to get started with, but the performance of my code on large circuits was an unexpected barrier later on, as well as the poor documentation of the different pseudorandom generators in TypeScript. Also, I was often too focused on quickly getting results rather than investing time in creating a convenient interface for the experiments or processing the resulting data, which led to wasted time in the long run.

@@

I started out by writing a program to reduce a single circuit to SAT using the Kojevnikov reduction, but quickly found it unwieldy and wrote a library of TypeScript objects to make it easy to read truth tables from files, store CNFs, quickly add certain types of clauses to them, and output them in the DIMACS format used by the SAT solvers.

>>>>

\begin{titlepage}
\title{Searching for circuit complexity lower bounds using SAT solvers}
\author{Candidate 1042155\\Trinity Term 2022\\MCompsci Computer Science}
\maketitle
\end{titlepage}

\begin{abstract}
Where the abstract will go
\end{abstract},

@@

A Boolean circuit consists of a network of gates computing simple boolean functions such as AND, OR and NOT, connected such that on any input, the gates compute some set of outputs. Finding lower bounds on the size of circuits computing given Boolean functions is one of the most fundamental and most challenging problems in computer science. Results in this area can be used to separate computational complexity classes.

For example, since P (the class of problems decidable in polynomial time in the length of the input) is contained within $P_{/poly}$ (the class of problems decidable by a family of circuits each of whose size is polynomial in the length of the input), then to separate P from NP (problems decidable by a non-deterministic Turing machine in polynomial time) it suffices to prove that NP $\notin P_{/poly}$, i.e. to find a function in NP which cannot be decided by a polynomial-size family of circuits. This would prove that a large class of fundamental and practically useful problems do not necessarily have polynomially fast algorithms~\cite{arora}.

The majority of functions do require exponential circuits, which can be seen from a simple counting argument noting that the number of possible circuits is far higher than the number of possible Boolean encodings of small circuits~\cite{arora}. Despite this fact, the best lower bounds proven on unrestricted circuits are only linear~\cite{boppana} and we still do not have optimal circuits or close upper and lower bounds for many important functions.

If functions with circuits of a given complexity could be identified by an efficient algorithm, this could be used to construct efficient learning algorithms for circuits of a smaller complexity; this would resolve one of the biggest questions in learning theory~\cite{carmosino}. Furthermore, such an algorithm could be used to break pseudorandom generators (i.e. distinguish random from pseudorandom inputs)~\cite{razborov94}. As a result, this problem has strong connections in both learning theory and cryptography.

A heuristic approach is to find minimal circuits for small instances of the functions. This can be used to prove new bounds~\cite{kulikovsurvey} and may lead to theoretical insights on the structure of their optimal circuits generally~\cite{williams}. Doing this for small functions computing natural like parity, modulo and addition allows us to gain a greater understanding of their difficulty and compute them more efficiently when using low-level code or hardware.

Finding efficient circuits is also necessary in practice when designing electronic systems. Electronic circuit design is an extensive field of its own, involving circuit simulation software to test designs, or hardware specification languages such as Verilog and circuit synthesis software to convert specifications into circuits. However, electronic circuit design is concerned with a wider range of problems (for example, functions whose output changes over time) and with concerns about correct physical implementation; we focus on abstract circuits.

One approach to automating the search for efficient circuits is to reduce the problem of designing a correct circuit (logical design synthesis) to the Boolean satisfiability problem (SAT). Existing algorithms (SAT solvers) can then be used to solve the problem, either finding a correct circuit of fixed size or generating a proof that none exists~\cite{kulikov}. SAT solvers have become increasingly powerful in recent years and now have many industrial applications\cite{prasad}; the wide range of heuristics implemented by different SAT solvers means that one suited to the problem at hand could be found or developed.

@@

Circuits are a popular model of computation; they are non-uniform, meaning that in a family of circuits where each takes a different size of input, each circuit is independent and not constructed by a common algorithm. Notably, this means that circuits are not subject to the Space and Time Hierachy Theorems, although there is a Non-Uniform Hierarchy Theorem stating that larger circuits can compute strictly more functions than smaller ones~\cite{arora}. In the past there have been theoretical proofs of lower bounds for restricted circuit classes, such as $AC^0$ (functions computed by circuit families of $O(1)$ depth)~\cite{furst81}~\cite{ajtai83}, $AC^0[k]$ (i.e. $AC^0$ additionally with MOD-k gates)~\cite{razborov87}~\cite{smolensky87}, and monotone circuits~\cite{razborov85}. There was also a bound showing the practical impossibility of deciding bounded-size formulae in a certain second-order language~cite{stockmeyer02}. However, no lower bounds better than (3+1/86)n-o(n) have been found for general circuits over the basis $B_2$ consisting of all functions on 2 bits~\cite{find16}, while the best bound is 5n-c for the basis $U_2 = \{ \land, \lor, \neg\}$~\cite{lachish01}~\cite{iwama02}. Furthermore, there are results stating that it is impossible to use circuit lower bounds to separate P and NP by relativising methods~\cite{baker75}, algebrizing methods~\cite{aaronson09}, or (under standard cryptographic assumptions) so-called `natural proofs'~\cite{razborov94}. A formula in Disjunctive Normal Form (DNF) is one consisting of clauses (disjunctions of literals), with the final formula consisting of the conjunction of all clauses; a k-DNF is one where the number of literals in one clause is at most k. It has been proved that circuit lower bounds are hard to prove using resolution and resolution with k-DNFS, and it is conjectured that they are also hard for stronger proof systems such as Frege systems~\cite{raz}.

Due to this difficulty in theoretical progress, we would like to investigate the problem empirically. SAT solvers provide a useful heuristic, as they allow us to examine small examples of the problem and find concrete rather than asymptotic bounds; due to improvements in SAT solver performance over the past years, this approach has now been put into practice. For example, Kamath et al~\cite{kamath} propose a reduction from logical design synthesis to SAT which fixes the DNF structure of the circuit and number of disjunctions used. Experiments using this reduction to find minimal circuits for 2-4 bit adders and multipliers are reported in~\cite{estrada}. Kojevnikov et al~\cite{kulikov} demonstrate a more general reduction allowing any circuit with a fixed number of gates, proving linear bounds for \(MOD^n_3\) circuits (computing whether the number of bits is equal to n modul0 3) over different Boolean bases. This was done by checking for circuit `building blocks' with 5 to 11 gates. In~\cite{kulikovlocal} Kulikov et al. used the same reduction to improve large circuit building blocks by searching for smaller versions of their sub-circuits. Specifically, linear upper bounds were proved or verified for \(SUM\), \(MOD^n_4\), and \(MOD^n_3\) by checking for circuits with up to 13 gates (they remarked that checking for existence of a Boolean circuit of size 13 was out of reach). Knuth also used SAT solvers to investigate lower bounds for \(MOD^n_3\)~\cite{knuth15}.

@@

We also search for a general relationship between solver performance and minimum circuit size by applying our methods to a sample of randomly generated functions. We then use solvers to prove a result, that a lower bound conjecture about MOD-3 functions due to Knuth~\cite{knuth15} does not hold for the de Morgan basis $U_2$.

@@

Denote by \(B_{n,m}\) the set of all Boolean functions \(f: \{0,1\}^n \to \{0,1\}^m\) and let \(B_n = B_{n,1}\). A circuit over the basis \(A \subseteq B_2\) is a finite directed acyclic graph with nodes of in-degree 0 or 2. Nodes of in-degree 0 are called inputs. Nodes of in-degree 2 are assigned functions from \(A\) and are called gates. Some nodes may be also be marked as outputs.

Without loss of generality, we can also assert that every gate has a larger index than both of its predecessors (i.e. the gates are sorted topologically w.r.t. the used numbering); that every gate's second predecessor has a larger index than its first predecessor; and that the last gate is an output ~\cite{kulikov}.

A circuit \(C\) of size $N$ is said to compute the function \(f\in B_{n,m}\) if, for all input vectors \(v\in\{0,1\}^n\) in the range of \(f\), \(C(v)=f(v)\), where \(C(v) = (o_1,...o_m)\) are the values of the output gates of \(C\), each input \(x_i\) takes value \(v_i\), and each gate takes the value of its assigned function when applied to the value of its two predecessor nodes.

In this paper, we mainly consider circuits over the De Morgan basis \[U_2 = \{\lor, \land, \neg\}\] where \(\neg\) denotes the Boolean function which outputs the negation of its first argument.

The size of a circuit is its number of gates. We define the Minimum Circuit Size Problem (MCSP) as the question of whether, given a natural number $N$ and a function \(f: \{0,1\}^n \to \{0,1\}^m\), there exists a circuit of size $N$ computing \(f\).

@@

SAT solvers have previously seen success in proving mathematical theorems, most famously in 2016 when Heule et al solved the Boolean Pythagorean Triples problem~\cite{heule} and in 2020 when Brakensiek, Heule et al solved Keller's conjecture~\cite{heule2}. In general, SAT solvers have the advantage of being effective even when heuristics are not known for solving the original problem, as in the case of MCSP.

We used three SAT solvers, all based on conflict-driven clause learning (CDCL). This is an algorithm closely related to the Davis–Putnam–Logemann–Loveland (DPLL) algorithm, and which is also based on the resolution proof system~\cite{cdcl}~\cite{krajicek}. As mentioned earlier, this means that they cannot decide general circuit lower bounds with a less than exponential-length proof~\cite{raz}. However, this asymptotic bound still allows for reasonably-sized proofs for small inputs such as the ones investigated in this paper.

The general idea of CDCL is to choose an assignment to a variable, propagate it (simplify the problem by removing clauses that now evaluate to true), and repeat until either finding a satisfying assignment, or finding that our assignment makes the formula unsatisfiable: in which case, add a `learned clause' to the formula according to a procedure incorporating resolution, and then `backtrack' (remove recent assignments so the formula is no longer unsatisfiable - if this is impossible, the answer is UNSAT) before trying again.

MiniSAT~\cite{minisat} was the winner of the 2005 SAT Competition and was last updated to version 2.2 in 2010~\cite{minisat2010}. Despite being old, it is well-documented and the other solvers we used were based on it, making it a good baseline for performance. It was also one of the solvers used to decide MCSP in~\cite{kulikov}.

PicoSAT is a solver inspired by miniSAT 1.14, with a focus on improving low-level performance by using optimised data structures~\cite{picosat}. It was submitted in the 2010 SAT-Race, and was also one of the solvers used in~\cite{kulikov} and the main one in~\cite{kulikovlocal}.

@@

Finally, in order to make full use of the Oxford ARC service and its distributed computing, the parallel solver Glucose-Syrup was used. This is a 2014 parallel version of the Glucose solver, which placed 4th in both the 2014 and 2015 SAT Competition. Glucose is a solver also based on MiniSAT 2.2, which uses a new heuristic to more accurately estimate the 'usefulness' of its learned clauses and regularly delete ones that are not as useful~\cite{glucose}. Glucose-Syrup is an unusual 'non-portfolio' parallel solver which runs multiple instances of the same solver in parallel, but rather than taking a divide and conquer approach, uses new heuristics for sharing learned clauses between the instances in order to speed them up~\cite{glucose-syrup}.

@@

\begin{enumerate}
  \item Let $l_1$, $l_2$ be existing literals chosen from the formula and let $q$ be a fresh variable, known as the extension variable.
  \item Add the clauses
  \[
    \{\neg l_1, q\}, \{\neg l_2, q\}, \{l_1, l_2, \neg q\}
  \]
  These are known as extension axioms.
  \item Repeat arbitrarily many times with new choices for $l_1$, $l_2$ and new fresh variable $q$. Note that the previously added $q$ may also be chosen as a literal.
\end{enumerate}

These axioms allow the SAT solver to use `Extended Resolution' - resolving on a formula incorporating multiple literals, rather than on a single literal. This is possible since $q \leftrightarrow l_1 \lr l_2$ and $l_1$, $l_2$ may also be extension variables, allowing for larger formulae to be built~\cite{krajicek}.

@@

We encoded the MCSP as a Boolean formula using three reductions. Let $t$ be the target function with $n$ inputs, and let $N$ be the number of gates required in the circuit.

@@

The first is due to Kojevnikov et al~\cite{kulikov} who present it together with bounds on the growth rate of the formula. In particular, the growth rate of the number of clauses is $O(N \cdot 2^n)$. The basis of the circuit can be restricted to any subset of \(B_2\).

@@

The second is due to Razborov~\cite{raz}, assumes the function outputs only a single Boolean value, and requires a circuit over the basis \(U_2\). Note that it has a lower growth rate (in terms of number of clauses) than the other two reductions - $O(N \cdot 2^n)$ as opposed to $O(N^3 \cdot 2^n)$ - suggesting it encodes the requirements more efficiently.

@@

The third reduction is straightforward. It also assumes the function outputs only a single Boolean value and requires a circuit over the basis \(U_2\). The growth rate for the number of clauses is $O(N \cdot 2^n)$, the same as for the Kojevnikov reduction.

@@

Timed experiments were carried out on a 3.400GHz Intel i3-8130U CPU, on a computer running Linux. The times shown were obtained by running the software on the problem 3 times and taking the median to ensure no anomalous times.

Further experiments were conducted on the Oxford Advanced Research Computing (ARC) service, which provides access to high-powered computing. A 48 core Cascade Lake (Intel Xeon Platinum 8268 CPU @ 2.90GHz) CPU was used~\cite{arc}.

@@

We investigated the relationship between the size of a circuit and the time taken for a SAT solver to confirm or deny its existence. We encoded a range of simple symmetric functions using the Kojevnikov encoding:

@@

The size of the minimal circuit is shown in red when it is under 10 gates.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.75\textwidth]{images/times/8and_kulikov.png}
  \caption{Times taken by SAT solvers to check for existence of a circuit computing AND on 8 bits, using the Kojevnikov reduction\label{fig:8and_kulikov}}
  \includegraphics[width=0.75\textwidth]{images/times/4parity_naive.png}
  \caption{Times taken by SAT solvers to check for existence of a circuit computing parity on 4 bits, using the Naive reduction\label{fig:4parity_naive}}
\end{figure}

It was found that the time needed to solve a problem tended to spike on or within 1 gate of the optimal circuit size. This may be because it takes longer to confirm or rule out circuits that come close to satisfying the requirements. Another possibility is that when close to the optimal size, there is the lowest proportion of satisfying circuits relative to the total amount of possible circuits, meaning that more of the search space must be explored (as opposed to quickly exploring it all at small sizes or quickly landing upon a satisfying circuit at large sizes). A similar phenomenon was reported in~\cite{estrada}. However, the time needed also tended to increase with the size of the circuit, even after a spike, reflecting the reductions' exponential growth in problem length with respect to size.

Different functions took different amounts of time overall, varying between seconds and hours, something also seen in Section \ref{randomresults}, and as discussed there, the time appeared to be greater when the function had a greater optimal circuit size. Performance for each function was not especially consistent between the solvers; usually they found the same things difficult, but the length of time each solver took for the difficult problems varied, shown best in Figure \ref{fig:8and_kulikov}. The extent of this can be seen in Section \ref{satresults}.

@@

@@

We were interested in what would be the fastest solver and reduction to use for further experiments, so we investigated whether there was a relationship between the reduction used to encode a Boolean function, the solver used, and the time taken by the solver. We are not aware of any prior investigation into this area.

We used the same three reductions and the MiniSAT, PicoSAT, and MapleSAT solvers. Glucose-Syrup was not included since it is a parallel solver, meaning the amount of time it takes is subject to different factors, e.g. number of threads available from the hardware and parallelisability of the problem, so it is not comparable.

We took the total of the times each combination took to solve the problems from Section \ref{funcresults} and compared them as shown in Figure \ref{fig:totals}.

\begin{figure}[h!]
  \includegraphics[width=\textwidth]{images/totals.png}
  \caption{Comparison of times taken by SAT solvers on different reductions}
  \label{fig:totals}
\end{figure}

It can be seen that MapleSAT on the Kojevnikov encoding performs the best. Apart from this, the effect of the reduction and the solver seem to be independent.

MapleSAT consistently performs better than MiniSAT and PicoSAT (probably because it is the most recent and builds upon the research that went into the other two), and PicoSAT occasionally `spikes' far worse on difficult problems than the other two, suggesting something about its modifications makes it especially unsuited for solving these reductions.

The Naive and Kojevnikov reductions both performed better than the Razborov reduction, but which was best depended on the solver used. It is possible that the efficient encoding of the Razborov reduction is detrimental to the performance of the solver. However, these results are only applicable to small values of n.

\clearpage

@@

The results are displayed in Figures \ref{fig:extparity} and \ref{fig:extmod3} as boxplots, where each `box' contains half of the datapoints for that result, and the median is shown as a horizontal line dividing the box in half. The lines above and below each box (`whiskers') show the maximum and minimum, with all datapoints lying between them.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.75\textwidth]{images/extvars/boxplot_4parity_medians.png}
  \caption{Comparison of times taken to check existence of a 9-gate circuit for 4-bit parity  \label{fig:extparity}}

  \centering
  \includegraphics[width=0.75\textwidth]{images/extvars/boxplot_4mod3_medians.png}
  \caption{Comparison of times taken to check existence of a 9-gate circuit for \(MOD_3\)  \label{fig:extmod3}}
\end{figure}

@@

\clearpage

@@

\begin{figure}[!h]
  \includegraphics[width=\textwidth]{images/random/4bit_scatter.png}
  \caption{Time taken to check up to 11 gates for randomly chosen functions}
  \label{fig:4rand_scatter}
\end{figure}

The scatter graph shows a clustering in times between functions of the same optimal circuit size, but with a few outliers taking more time. This is most easily seen for sizes 10 and 11 (although the same is visible for 8 and 9 when plotted on a smaller scale). Also, the clustered functions of each optimal circuit size took strictly less time to solve than those with larger sizes, despite the fact that we always checked for existence up to 11 gates. On closer inspection of the times for each function, a spike in length on or around the minimum size was usually observed, similarly to in \ref{satresults}. This shows a generally strong relationship between the difficulty of checking circuit existence and the complexity of the function.

\begin{figure}[!h]
  \includegraphics[width=\textwidth]{images/random/4bit_sizes.png}
  \caption{Sizes of optimal circuits for 4-bit functions with randomly chosen truth tables}
  \label{fig:4rand_sizes}
\end{figure}

While we expect a large proportion of circuits to be hard (i.e. have a large optimal circuit size) for large input sizes ~\cite{arora}, the usual counting arguments break down for such small values of n. However, using the SAT solver we can check the number of hard circuits directly. While the small sample size can account for the irregular shape of the graph in \ref{fig:4rand_sizes}, there does appear to be a peak around 10 gates. This contradicts the intuition that there should be exponentially more harder functions, and suggests that there is quite a low upper limit on the complexity of a function which only can only vary over 4 bits of input. However, this is all under the assumption that the pseudorandom generator used did output a truly random sample; in reality, it may have sampled disproportionately many easy (or hard) functions.

@@

Define $MOD^{m,r}_n(x_1 \ldots x_n) = [x_1 + \ldots + x_n \equiv r $mod$ m]$, where $[b]$ is the Iverson bracket, i.e. evaluates to 1 if $b$ evaluates to true, else false.

@@

We decided to try and check the hypothesis for $n=6, r=0$ and $n=6, r=1$ for the De Morgan basis. This involved checking for circuits for the function up to size 11 and 12, and 12 and 13 respectively. The Glucose-Syrup solver was used, running on the Oxford ARC service with a maximum of 32 threads (so, 32 parallel instances of the server working together). In order to ensrue that the number of threads would lead to a gain in performance, rather than a drop due to the extra time needed to coordinate them, the SAT solver was timed on a smaller problem with 4, 8, 16 and 32 threads; it displayed gains in performance up to 32.

@@

@@

The main practical lesson learned from this project was to think in the long term. Firstly, by writing code that can easily be adapted, i.e. following good OOP practices, so that less refactoring will be needed later, and by writing code that can scale to dealing with larger numbers as the scope and ambition of the project expands. This could be done by choosing a well-suited programming language, algorithms and data structures - I chose TypeScript (a statically typed variant of Javascript) and used arrays to store my CNF formulae, since these were familiar and easy for me to get started with, but the performance of my code on large circuits was an unexpected barrier later on, as well as the poor documentation of the different pseudorandom generators in TypeScript. Also, I was often too focused on quickly getting results rather than investing time in creating a convenient interface for the experiments or processing the resulting data, which led to wasted time in the long run.

@@

I have begun to get a feel for how SAT solvers can be used in theoretical Computer Science. I have observed some of the current capabilities of solvers, the categories under which they are judged (such as general, industrial, and parallel) and the algorithms which power them, including by coding my own basic DPLL algorithm during preliminary experiments. I have learned more about the current state of research in circuit complexity and learned the idea of proof complexity as a result of my background reading, as well as discovering the various barriers that actually make proving the P vs NP problem so difficult.
